Efficient Feature Screening for Lasso-Type Problems via Batched Safe-Strong Rules

### 2020-01-22 main.tex
Fixed error (missing $ for math); compiles properly now

\note{I think we need a better transition here} In this paper, we proposed an adaptive batched screening scheme that changes how screening is done along the path of $\lambda$ values and implement it into hybrid approaches to form adaptive batched safe-strong rules. We divided the set of $\lambda$ values into batches and screening within each batch is based on the same solution at one $\lambda$ value at the beginning of the batch. On one hand, basing on the same solution avoids many costly computation and on the other hand, referring to solution at a close $\lambda$ value provides more screening power. An adaptive choice of batches sizes are applied using the screening performance from previous $\lambda$ values to better balance these two effects. It will be shown that our adaptive rule can greatly reduce the computation time and outperform other method under all scenarios. Similar to hybrid approaches, our method can also utilize any member of safe rules, giving it the potential for further speed up given new safe rules and the flexibility to extend to other lasso type models. For example we extended the method to work with logistic regression with lasso penalty.

@@

In fact, the solution of the primal problem (1) and the solution of the dual problem (7) can be linked by:
\begin{equation}
    \hat{\theta}(\lambda)=\frac{y-X\hat{\beta}(\lambda)}{n\lambda}=\frac{r(\lambda)}{n\lambda},
\end{equation}
so bounding $\hat{\theta}(\lambda)$ is essentially bounding the residual vector $r(\lambda)$. Then using the KKT conditions, some features can be guaranteed to have zero coefficients and be discarded safely. Two forms of the EDPP rule, basic EDPP (BEDPP) and sequential EDPP (SEDPP), were proposed\cite{wang2013lasso} simplified\cite{zeng2017efficient} under the standardization (3) and can be stated in the following theorems:

@@

Batch screening iterative lasso (BASIL) \cite{qian2019fast} was proposed to do SSR screening and its following post-convergence check in batches. Suppose we have a previous solution $\hat{\beta}(\lambda_k)$ and $r(\lambda_k)$ and the batch of $\{\lambda_{k+1},\lambda_{k+2},...,\lambda_{k+l}\}$ values we are going to solve at. First we do the screening for each $\lambda_{k+i}$ value in the batch based on solution at $\lambda_k$ by discarding the $j$-th feature if $|x_j^Tr(\lambda_k)/n|<2\lambda_{k+i}-\lambda_k$. Then at each $\lambda_{k+i}$ value we solve the lasso problem with only the features that are not discarded. Last at post-convergence check, we search for largest $k+i$ in the batch $\{k+1,k+2,...,k+l\}$ such that KKT conditions hold for solution at $\lambda_{k+i}$ and accept only the solution from $\lambda_{k+1}$ to $\lambda_{k+i}$. Next, we move to a new batch by replacing the batch head $k$ with $k+i$ and repeat the same procedure again.

The BASIL is most beneficial when the whole $X$ matrix is to large to be fit into the RAM, thus we need to read in a small number of columns $x_j$ from the hard disk at a time. In the post-convergence check for every batch, for each $j$, $x_j$ only needs to be read in once and can be used multiple times for $\{\lambda_{k+1},\lambda_{k+2},...,\lambda_{k+l}\}$. Moreover, similar to SSR with ordinary post-convergence check, $x_j^Tr(\lambda_{x+i})$ computed at BASIL's post-convergence check step can be reused for screening at next batch.

@@

\subsection{Batched EDPP rule}
\begin{itemize}
    \item Validate that given solution at one lambda value, we can safely screen for solutions at a batch of lambda values.
    \item From the simplified notation of the EDPP rule, most quantities need to be calculated once per batch from the given solution.
    \item Propose the new rule.
\end{itemize}

\subsection{Batched strong rule after batched EDPP rule}
\begin{itemize}
    \item Analyse hybrid strong rule in this case and point out calculation that can be avoided (at a small cost of power)
    \item Propose the new combined rule (Batched safe strong rule)
    \item Proof of convergence
\end{itemize}

@@

\begin{itemize}
    \item Quantities that can be shared by batched safe strong rule and coordinate descent
    \item pseudo code
\end{itemize}

>>>>

Previous screening rules can also be divided into two categories, sequential rules and one-step rules, by how the screening is carried out along the path of $\lambda$ values. Sequential rules do heavy computation for each $\lambda$ values, so they are slower but relatively powerful at each $\lambda$ values. One-step rules only require heavy computation at the beginning and thus are faster but they cannot remain powerful along the path for long.

In this paper, we proposed an adaptive batched screening scheme that changes how screening is done along the path of $\lambda$ values and implement it into hybrid approaches to form adaptive batched safe-strong rules. We divided the set of $\lambda$ values into batches and screening within each batch is based on the same solution at one $\lambda$ value at the beginning of the batch. This screening scheme helps avoid disadvantages of both sequential rules and one-step rules. On one hand, basing on the same solution avoids many costly computation within a batch and on the other hand, referring to solution at a close $\lambda$ value provides more screening power. An adaptive choice of batches sizes are applied using the screening performance from previous $\lambda$ values to better balance these two effects. It will be shown that our adaptive rule can greatly reduce the computation time and outperform other methods under all scenarios. Similar to hybrid approaches, our method can also utilize any member of safe rules, giving it the potential for further speed up given new safe rules and the flexibility to extend to other lasso type models. For example we extended the method to work with logistic regression with lasso penalty.

@@

In fact, the solution of the primal problem (1) and the solution of the dual problem (7) can be linked by:
\begin{equation}
    \hat{\theta}(\lambda)=\frac{y-X\hat{\beta}(\lambda)}{n\lambda}=\frac{r(\lambda)}{n\lambda},
\end{equation}
so bounding $\hat{\theta}(\lambda)$ is essentially bounding the residual vector $r(\lambda)$. Then using the KKT conditions, some features can be guaranteed to have zero coefficients and be discarded safely. Two forms of the EDPP rule, basic EDPP (BEDPP) and sequential EDPP (SEDPP)\cite{wang2013lasso}, were proposed and under the standardization (3) can be simplified in the following theorems\cite{zeng2017efficient}:

@@

Batch screening iterative lasso (BASIL) \cite{qian2019fast} was proposed to do SSR screening and its following post-convergence check in batches. Suppose we have a previous solution $\hat{\beta}(\lambda_k)$ and $r(\lambda_k)$ and the batch of $\{\lambda_{k+1},\lambda_{k+2},...,\lambda_{k+B}\}$ values we are going to solve at. First we do the screening for each $\lambda_{k+b}$ value in the batch based on solution at $\lambda_k$ by discarding the $j$-th feature if $|x_j^Tr(\lambda_k)/n|<2\lambda_{k+b}-\lambda_k$. Then at each $\lambda_{k+b}$ value we solve the lasso problem with only the features that are not discarded. Last at post-convergence check, we search for largest $k+b$ in the batch $\{k+1,k+2,...,k+B\}$ such that KKT conditions hold for solution at $\lambda_{k+b}$ and accept only the solution from $\lambda_{k+1}$ to $\lambda_{k+b}$. Next, we move to a new batch by replacing the batch head $k$ with $k+b$ and repeat the same procedure again.

The BASIL is most beneficial when the whole $X$ matrix is to large to be fit into the RAM, thus we need to read in a small number of columns $x_j$ from the hard disk at a time. In the post-convergence check for every batch, for each $j$, $x_j$ only needs to be read in once and can be used multiple times for $\{\lambda_{k+1},\lambda_{k+2},...,\lambda_{k+B}\}$. Moreover, similar to SSR with ordinary post-convergence check, $x_j^Tr(\lambda_{x+b})$ computed at BASIL's post-convergence check step can be reused for screening at next batch.

@@

In this section, we will introduce the batched safe-strong rules, combine them with pathwise coordinate descent algorithm and propose a strategy to adaptively  start new batches.

\subsection{Batched EDPP rule}

From the SEDPP rule (10) in Theorem 2.2, we can do screening at $\lambda_{k+1}$ given the solution at $\lambda_k$. Now consider we are solving the lasso problem along the path $\lambda_0,...,\lambda_k,\lambda_{k+2},...,\lambda_K$ instead. Using the SEDPP rule, we can do screening at $\lambda_{k+2}$ as well, given the solution at $\lambda_k$, which is considered the previous solution now. Similarly, given the solution at $\lambda_k$, we can do screening for a batch of $B$ $\lambda$-values: $\lambda_{k+1},...,\lambda_{k+B}$ and then all the discarded features are guaranteed to be zero. Batched EDPP is still a safe rule, but it is neither sequential nor one-step.

Similar to SEDPP, the batched EDPP rule can be stated in the following theorem:

\begin{theorem}[Batched EDPP]
    For the lasso problem (1), let $\lambda_0:=\max_j|x_j^Ty/n|$. Suppose we have a sequence of $\lambda$ values $\lambda_0>\lambda_1>...>\lambda_K$. Under condition (3)
    \begin{enumerate}
        \item For any $k=1,2,...,K-1$, given $\hat{\beta}(\lambda_k)$, $r(\lambda_k)$ and $\hat{y}(\lambda_k):=y-X\hat{\beta}(\lambda_k)$, then for some $B$ and any $b$ s.t. $k<k+b\leq k+B\leq K$, $\hat{\beta}_j(\lambda_{k+b})=0$ if
        \begin{equation}
            \begin{split}
                &\left|2\lambda_{k+b}x_j^Tr(\lambda_k)+(\lambda_k-\lambda_{k+b})\left( x_j^Ty-\frac{y^T\hat{y}(\lambda_k)x_j^T\hat{y}(\lambda_k)}{||\hat{y}(\lambda_k)||_2^2}\right)\right|\\&<2n\lambda_k\lambda_{k+b}-(\lambda_k-\lambda_{k+b})\sqrt{n||y||_2^2-\frac{n(y^T\hat{y}(\lambda_k))^2}{||\hat{y}(\lambda_k)||_2^2}}
            \end{split}
        \end{equation}
        \item For k=0, i.e., $\lambda_k=\lambda_0$, then for some $B$ and any $b$ s.t. $0<b\leq B\leq K$, $\hat{\beta}_j(\lambda_{k+b})=0$ if
        \begin{equation}
        |(\lambda_0+\lambda_b)x_j^Ty-(\lambda_0-\lambda_b)sign(x_*^Ty)\lambda_0x_j^Tx_*|<2n\lambda_0\lambda_b-(\lambda_0-\lambda_b)\sqrt{n||y||_2^2-n^2\lambda_0^2}.
    \end{equation}
    \end{enumerate}
\end{theorem}

For each $\lambda$-value in the batch $\lambda_{k+1},...,\lambda_{k+B}$, the computation in (11) or (12) is the same, except for one quantity $\lambda_{k+b}$ ($\lambda_b$ in the case $k=0$). The most expensive computations are $x_j^Tr(\lambda_k)$ and $x_j^Ty$, for $j=1,2,...,p$. $x_j^Ty$ are the same for all $\lambda$-values, $x_j^Tr(\lambda_k)$ are the same for all $\lambda$-values within the batch and $x_j^T\hat{y}(\lambda_k)=x_j^Ty-x_j^Tr(\lambda_k)$, so once they are computed, they can be stored and reused to avoid more computation in that batch. These two computation take $O(np)$ time and the rest computation in (11) and (12) take at most a lower order of $O(p)$ time.

Compared to sequential methods like SEDPP, the batched method only requires one $O(np)$ computation in one batch, while SEDPP requires one $O(np)$ computation for each $\lambda$-value. Compared to one-step methods like BEDPP, if the batch size is not too large, the solution at the batch head $\lambda_k$ and the solution at some $\lambda_{k+b}$ in the batch are usually relatively close, so screening at $\lambda_{k+b}$ using solution at $\lambda_k$ will remain powerful. For one-step method, as $\lambda_k$ gets further away from $\lambda_0$, the screening power can drop drastically and
may not be any beneficial after that.

\subsection{Batched strong rule after batched EDPP rule}

Because HSSR \cite{zeng2017efficient} have been shown to provide significant speedup compared to safe rules alone, we should also consider to add a strong rule screening after the batched EDPP rule screening. The batched SSR and batched EDPP rule combined will form our batched safe strong rule.

For ordinary SSR, the quantities $x_j^Tr(\lambda_k)$ used in (4) is already computed in the previous post-convergence check in (2). However, combined with safe rules, previous post-convergence check only need to check features that are not discarded at the previous $\lambda$-value. Thus, for features that are discarded at the previous $\lambda$-value but not at the current $\lambda$-value, their corresponding $x_j^Tr(\lambda_k)$ have not been computed yet and thus cannot be used without additional cost. Similar to the idea of BASIL \cite{qian2019fast}, we can also do the strong rule screening in batches to reduce the number of required computations.

we are going to combine batched EDPP rule and batched SSR to form the batch safe strong rule as following. First, we determine a batch $\lambda_{k+1},...,\lambda_{k+B}$, and both batched EDPP rule and batched SSR will be conducted in this batch. Second $x_j^Tr(\lambda_k)$ will be computed and stored. Then for each $\lambda_{k+b}$ in the batch, after doing batched EDPP rule screening, a batched SSR screening is carried out on the features have not been discarded by batched EDPP rule and discard features that satisfies: $|x_j^Tr(\lambda_k)/n|<2\lambda_{k+b}-\lambda_k$. Last, a post-convergence check is conducted on the features that were discarded by batched SSR but not discarded by batched EDPP rule.

By choosing the same batch for batched EDPP rule and batched SSR, the quantity $x_j^Tr(\lambda_k)$ only needs to be computed once and can be used by both rules. Also, because batched EDPP rule can stay powerful, it keeps the set of features that require post-convergence check small, while because basic EDPP loses power for $\lambda$-values that are far away from $\lambda_0$, post-convergence check for HSSR will be more expensive.

@@

In this section, we will show the algorithm to do pathwise coordinate descent with our batched safe strong rule screening. We will only show the algorithm for solving one batch $\lambda_{k+1},...,\lambda_{k+B}$. To solve the next batch, we just need to apply the same algorithm using the output from last batch as input. Note here for the first batch with batch head $\lambda_0$, the residuals vector $r(\lambda_0)$ is just the response $y$, because all the $\hat{\beta}(\lambda_0)$ are zero. $\{x_j^Ty\}_{j=1}^p$ need to be computed at the beginning and stored for the whole path.

\begin{algorithm}[H]
    \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Pre}{Pre-compute}\SetKwInOut{Initialize}{Initialize}
    \SetAlgoLined
    \Input{$\{x_j\}_{j=1}^p, \{x_j^Ty\}_{j=1}^p, \hat{\beta}(\lambda_k), r(\lambda_k), \lambda_{k+1},...,\lambda_{k+B}$}
    \Pre{ $\{x_j^Tr(\lambda_k)\}_{j=1}^p, \{x_j^T\hat{y}(\lambda_k)=x_j^Ty-x_j^Tr(\lambda_k)\}_{j=1}^p, \hat{y}(\lambda_k)=y-r(\lambda_k)$}
    \BlankLine
    \For{$b\xleftarrow{}$ 1 \KwTo B}{
        Safe Screening: $\mathcal{S}\xleftarrow{}\{j:x_j$ not discarded by batched EDPP rule$\}$.

        Strong Screening: $\mathcal{R}\xleftarrow{}\{j\in \mathcal{S}:x_j$ not discarded by batched SSR$\}$.

        \Initialize{$\hat{\beta}(\lambda_{k+b})=\hat{\beta}(\lambda_{k+b-1}), r(\lambda_{k+b})=r(\lambda_{k+b-1})$}

        \Repeat{Post-convergence check passes for all $j\in \mathcal{S}\setminus\mathcal{R}$}{
            Conduct pathwise coordinate descent in Algorithm 1 with warm start $\hat{\beta}(\lambda_{k+b}), r(\lambda_{k+b})$, until converge. Save the outputs as $\hat{\beta}(\lambda_{k+b}), r(\lambda_{k+b})$.
        }
    }
    \Output{$\{\hat{\beta}(\lambda_{k+b})\}_{b=1}^B, r(\lambda_{k+B})$}
    \caption{Pathwise coordinate descent with batched safe strong rule screening}
\end{algorithm}
